---
title: "Getting started with Ipse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Ipse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Ipse)
```

## What this package does

`Ipse` assigns a cell type label to each cell (or spot) in an
expression matrix, using a marker gene reference bundled inside the
package — no external atlas download required. This vignette walks through
the basic workflow on a small simulated dataset; swap in your own
`expr` matrix (genes x cells, gene symbols as row names) to use it for
real.

## A toy example

```{r toy-data}
set.seed(1)
genes <- unique(midbrain_markers$gene)
n_cells <- 20

expr <- matrix(
  rpois(length(genes) * n_cells, lambda = 3),
  nrow = length(genes),
  dimnames = list(genes, paste0("cell", seq_len(n_cells)))
)

# Simulate five clearly dopaminergic cells by boosting their markers
da_genes <- midbrain_markers$gene[midbrain_markers$cell_type == "Dopaminergic neuron"]
expr[da_genes, 1:5] <- expr[da_genes, 1:5] + 15
```

## Annotating cells

```{r annotate}
labels <- annotate_cells(expr)
table(labels$cell_type)
head(labels)
```

## Looking at raw scores

If you want the full cell-types x cells score matrix rather than hard
labels:

```{r scores}
scores <- score_markers(expr)
scores[, 1:5]
```

## What cell types are covered?

```{r list-types}
list_cell_types()
```

## Next steps

- Replace `data-raw/prepare_markers.R` with a curated marker list for your
  use case
- Try `annotate_cells(expr, min_score = ...)` to flag
  low-confidence calls as `"Unassigned"` instead of forcing every cell into
  a category
