---
title: "Choosing a scoring method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a scoring method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(Ipse)
```

## Why the method matters

`score_markers()` offers four ways to turn marker expression into a
per-cell-type score. On clean, well-separated data the choice barely
matters — every method recovers the right answer. The differences appear
when marker sets differ in **absolute abundance**, which is the normal
case: `MBP` and `PLP1` are enormously expressed, `SLC6A3` is not.

The four methods differ in what they normalise against:

| method | normalises against | corrects abundance imbalance? | cells scored independently? |
|---|---|---|---|
| `mean_weighted` | nothing | no | yes |
| `control` | expression-matched background genes | only additively — partially | yes |
| `ucell` | other genes *within the same cell* | no | yes |
| `zscore` | the same gene *across other cells* | yes | no |

## A benchmark

Two simulated scenarios, both with 2,000 background genes:

- **clean** — markers are near-silent outside their own cell type
- **spillover** — oligodendrocyte markers are constitutively abundant in
  *every* cell, and dopaminergic markers are scarce. This mimics ambient RNA
  and transcript spillover, which is a routine problem in imaging-based
  spatial assays.

Approximate accuracy against known ground truth:

| method | clean | spillover |
|---|---|---|
| `mean_weighted` | 100% | ~17% |
| `control` | 100% | ~17% |
| `ucell` | 100% | ~17% |
| `zscore` | 100% | ~46% |

Two things are worth taking from this.

**First, only `zscore` degrades gracefully.** The reason is structural: it
is the only method that compares a gene against its own distribution across
cells, which is what distinguishes "this gene is high here" from "this gene
is high everywhere". Background subtraction (`control`) removes an additive
offset but leaves the difference in dynamic range intact. Within-cell
ranking (`ucell`) is defeated by a marker set that ranks top in every cell.
This is why `zscore` is the default.

**Second, and more importantly — no method rescues the spillover scenario.**
Even the best falls below 50%. The scoring formula is not where this problem
is solved. Marker curation and upstream handling of spillover matter more
than the choice of score.

## When to override the default

Use `ucell` when you need scores that are comparable across separately
processed batches or tiles, or when you want to score cells in chunks —
`zscore` scores depend on which other cells are in the matrix, `ucell`
scores do not:

```{r independence, eval = FALSE}
full   <- score_markers(expr, method = "ucell")
subset <- score_markers(expr[, 1:10], method = "ucell")
all.equal(full[, 1:10], subset)   # TRUE for ucell, FALSE for zscore
```

Also prefer `ucell` when your dataset is dominated by a single cell type, a
case where z-scoring has little variation to work with.

## Flagging uncertain calls

Rather than forcing every cell into a category, use the two thresholds:

```{r thresholds, eval = FALSE}
labels <- annotate_cells(
  expr,
  min_score  = 0,    # cell must beat its background -> else "Unassigned"
  min_margin = 0.1   # top two types must be separated -> else "Ambiguous"
)
table(labels$cell_type)
```

Inspecting the `margin` column is often more informative than the label
itself: a large population of low-margin cells usually means the marker set
does not separate the types present in your data, which is a marker
problem, not a threshold problem.
